\documentclass[11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{mathtools}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{colorlinks=true, linkcolor=blue!50!black, urlcolor=blue!50!black}

\newcommand{\sat}[2]{[#1]\!:\!#2^{\infty}}
\newcommand{\satalg}[2]{(#1)\!:\!#2^{\infty}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\ld}{\operatorname{ld}}
\newcommand{\sep}{\operatorname{sep}}
\newcommand{\ini}{\operatorname{ini}}
\newcommand{\si}{\operatorname{si}}
\newcommand{\NF}{\operatorname{NF}}
\newcommand{\dZero}{\mathrm{d\text{-}Zero}}
\newcommand{\dprem}{\mathrm{d\text{-}prem}}
\newcommand{\R}{\mathfrak{R}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\UU}{\mathbb{U}}
\newcommand{\imp}{\;\Longrightarrow\;}

\title{Coherent, regular, simple, passive:\\
       a reconciliation of the two decomposition traditions\\
       {\large (Rosenfeld--Gr\"obner vs.\ differential Thomas)}}
\author{}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
\noindent
A companion note to \emph{A New Method of Solving PDEs}, recording the exact
relationship between four notions that recur whenever one compares the
Rosenfeld--Gr\"obner (RG) and differential-Thomas decompositions: a differential
triangular system may be \emph{coherent}, \emph{regular}, \emph{(differentially)
simple}, or \emph{passive} (\emph{= Janet-involutive}). These are \emph{not}
synonyms and they are \emph{not} mutually equivalent. They form a strict
one-directional hierarchy
\[
  \text{simple}\imp\text{regular}\imp\text{coherent},
  \qquad\qquad
  \text{passive}\imp\text{coherent},
\]
with coherence as the common base property (``all $\Delta$-polynomials reduce to
zero'') that makes Rosenfeld's Lemma fire. The note pins each node to a primary
source, quotes the defining conditions verbatim, and flags one correction to an
earlier informal claim (that Seiler's monograph states a coherence
$\Leftrightarrow$ involution \emph{equivalence}): the sourced statement is the
\emph{implication} passive $\Rightarrow$ coherent (Li--Wang 1999, Thm.~3;
Gerdt 1999, Thm.~16; B\"achler et al.\ 2012).
\end{abstract}

\section{Why this note exists}

RG and differential Thomas both decompose the radical differential ideal
$\{\PP\}$ generated by a finite set $\PP$ of differential polynomials, but they
return components of different \emph{kinds}:
\begin{itemize}
  \item RG returns \emph{regular} differential systems and represents
        $\{\PP\}$ as an \emph{intersection} (components may overlap);
  \item differential Thomas returns \emph{simple} differential systems and
        gives a \emph{disjoint} decomposition (the solution sets partition).
\end{itemize}
Underneath both sits a third notion, \emph{coherence} (the condition in
Rosenfeld's Lemma), and on the Janet--Riquier side a fourth, \emph{passivity /
involution}. In loose usage these four words get treated as interchangeable.
They are not. The purpose of this note is to fix the definitions from primary
sources and record the exact implication lattice, so the distinction need not be
re-derived. The single most on-point reference is Li \& Wang (1999), whose very
title is the question --- ``\emph{Coherent, regular and simple systems in zero
decompositions of partial differential systems}'' --- and whose abstract states
its purpose as: \emph{``Rosenfeld's lemma is extended to a more general setting;
relations between passivity and coherence are clarified; regular systems and
simple systems are generalized and proposed.''}

\section{Standing notation}

Fix a differential field $K$ of characteristic $0$ with commuting derivations,
and the differential polynomial ring $\R=K\{X_1,\dots,X_n\}$. For a differential
polynomial $P\in\R\setminus K$: $\ld(P)$ is its \emph{lead} (highest-ranked
derivative under a fixed admissible ranking), $\ini(P)$ its \emph{initial}
(leading coefficient in $\ld(P)$), and $\sep(P)=\partial P/\partial\ld(P)$ its
\emph{separant}. Write $\si(\PP)$ for the set of separants and initials of
$\PP$, and $[\PP]$, $\{\PP\}$ for the differential and radical differential
ideals generated by $\PP$. Saturation: $\sat{\PP}{\,\si(\PP)}$ denotes the
differential ideal quotient by the multiplicative set generated by $\si(\PP)$.
$\dZero(\PP/\mathbb{Q})$ is the differential zero set of $\PP$ off the zero set
of $\mathbb{Q}$; $\dprem$ is the differential pseudo-remainder.

We use ``d-tri set'' for a differentially triangular set (distinct leads,
pairwise partially reduced) and ``d-asc set'' for a differential ascending
(autoreduced) set. Every d-asc set is a d-tri set; the converse fails, and Li \&
Wang work with the weaker d-tri notion throughout.

\section{The four definitions, from the sources}

\subsection{Coherence (Li--Wang 1999, Def.~5)}

The $\Delta$-polynomial of $F,G\in\R\setminus K$ of the same class is the
differential analogue of an $S$-polynomial (construction due to Rosenfeld):
\begin{equation}\label{eq:delta}
  \Delta(F,G)=\sep(G)\,\phi_1 F-\sep(F)\,\phi_2 G,
  \qquad \ld(\phi_1 F)=\ld(\phi_2 G),
\end{equation}
where $\phi_1,\phi_2$ are proper derivative operators of \emph{lowest} order
bringing the two leads to a common derivative.

\begin{quote}
\textbf{Definition (coherent).} A weakly-reduced set $\PP$ is \emph{coherent}
if, for any $F,G\in\PP$ of the same class and any $\Delta(F,G)$ as in
\eqref{eq:delta},
\[
  J\,\Delta(F,G)=C_1\theta_1 P_1+\cdots+C_r\theta_r P_r,
  \qquad \theta_i P_i<\ld(\phi_1 F),
\]
for some SI-product $J$, coefficients $C_i\in\R$, and $P_i\in\PP$.
\end{quote}

In words: \emph{every $\Delta$-polynomial reduces to a combination of
lower-lead derivatives of members of $\PP$}. This is precisely the differential
Buchberger criterion. (``Weakly reduced'' (Def.~3) means every $\ini(P)$ and
$\sep(P)$ is partially reduced w.r.t.\ $\PP$; every d-tri set is weakly reduced.)

\paragraph{What coherence buys: the generalized Rosenfeld Lemma.}
\emph{Li--Wang Theorem 2.} If $\PP\subset\R\setminus K$ is weakly reduced and
coherent, and $F\in\sat{\PP}{\,\si(\PP)}$ is partially reduced w.r.t.\ $\PP$,
then $F\in\satalg{\PP}{\,\si(\PP)}$. That is: coherence is exactly the
hypothesis under which \emph{differential} ideal membership collapses to
\emph{algebraic} ideal membership. This is the theoretical engine of the entire
RG algorithm (Boulier--Lazard--Ollivier--Petitot).

\subsection{Regular (Li--Wang 1999, Def.~6)}

First, the system-level upgrade of coherence: a d-tri system $\langle
\TT,\UU\rangle$ is \emph{coherent} if $\TT$ is coherent and each d-pol in $\UU$
is partially reduced w.r.t.\ $\TT$. Then:

\begin{quote}
\textbf{Definition (regular).} A d-tri system $\langle\TT,\UU\rangle$ is
\emph{regular} if it is a coherent \emph{and} elementarily-triangular (e-tri)
system.
\[
  \boxed{\ \text{regular}\ =\ \text{coherent}\ +\ \text{e-tri}\ }
\]
\end{quote}

Here $\langle\TT,\UU\rangle$ is \emph{e-tri} (Def.~1) if \emph{every} point in
$\mathrm{Zero}(\TT/\UU)$ --- the \emph{algebraic}, not just differential, zero
set --- fails to annihilate any element of $\si(\TT)$; equivalently the initials
and separants are invertible on the whole (algebraic) component, not merely
generically. Li \& Wang note this \emph{generalizes} Boulier et al.'s regular
chains in two ways: $\TT$ need not be a d-asc set, and $\si(\TT)$ need not be
contained in $\UU$.

\paragraph{What regularity buys.} \emph{Li--Wang Theorem 8.} For a regular
$\langle\TT,\UU\rangle$: a d-pol $F$ lies in $\sat{\TT}{\,\UU}$ iff a partial
pseudo-remainder of $F$ w.r.t.\ $\TT$ lies in $\satalg{\TT}{\,\UU}$; and both
$\sat{\TT}{\,\UU}$ and $\satalg{\TT}{\,\UU}$ are \emph{radical}. Regular systems
are the components of the RG ``weak form''
$\dZero(\PP/\mathbb{Q})=\bigcup_i\dZero(\TT_i/\UU_i)$ --- an
\emph{intersection}-style decomposition whose components can overlap.

\subsection{Differentially simple (Li--Wang 1999, Def.~8; algebraic core Def.~2)}

The algebraic prototype (Def.~2, after Thomas and Wang): a triangular system
$\langle\TT,\UU\rangle$ in $K[\mathbf x]$ is \emph{w-simple} if for each
$1\le k\le n$ it is triangular, has non-vanishing initials on the lower solution
set, and is \emph{square-free} (the univariate image $T(\bar{\mathbf
x}^{\{p-1\}},x_p)$ is square-free). Lifting to the differential setting:

\begin{quote}
\textbf{Definition (differentially simple, d-sim).} A \emph{coherent} d-tri
system $\langle\TT,\UU\rangle$ in $\R$ is \emph{differentially simple} if, when
considered as a triangular system in the underlying (non-differential) ring, it
is w-simple.
\end{quote}

So, informally,
\[
  \text{simple}\ =\ \text{regular}\ +\ \text{square-free / w-simple conditions}.
\]
The disjoint decomposition $\dZero(\PP/\mathbb{Q})=\bigcup_i\dZero(\TT_i/\UU_i)$
into d-sim systems is the (differential) \emph{Thomas decomposition}; its
components \emph{partition} the solution set.

\paragraph{What simplicity buys.} \emph{Li--Wang Theorem 9 \& Corollary 2.} A
d-sim system has $\dZero(\TT/\UU)\ne\emptyset$, and for any d-pol $P$,
\[
  \dZero(\TT/\UU)\subset\dZero(P)\iff\dprem(P,\TT)=0 .
\]
Moreover $\TT$ is then a characteristic set (in Ritt's sense) of the prime
differential ideal $\sat{\TT}{\,\UU}$. And, top of Li--Wang p.~58: ``\emph{a
d-sim system must be regular}.''

\subsection{Passive / involutive (Gerdt 1999; used by differential Thomas)}

``Passive'' is the Janet--Riquier word for the same completeness idea, phrased
via an \emph{involutive division} (Janet division for the Thomas construction).
Gerdt states the terminology bridge in prose: \emph{``A system satisfying the
Janet involutivity conditions is often called \textbf{passive}.''} The condition
itself:

\begin{quote}
\textbf{Definition (involutive, Gerdt 1999 Def.~14).} An $L$-autoreduced set $F$
is \emph{$L$-involutive} if
\[
  (\forall f\in F)\,(\forall\alpha\in\mathbb{N}^n)\;
  \bigl[\,\NF_L(\partial_\alpha f,\,F)=0\,\bigr].
\]
\textbf{Local criterion (Gerdt 1999 Thm.~16).} $F$ is $L$-involutive
(for a continuous involutive division $L$) iff
\[
  (\forall f\in F)\,(\forall x_i\in\mathrm{NM}_L(f,F))\;
  \bigl[\,\NF_L(\partial_{x_i}\!\cdot f,\,F)=0\,\bigr],
\]
i.e.\ only the \emph{non-multiplicative} prolongations need reduce to zero.
\end{quote}

This finite, checkable criterion is ``the involutivity criterion'' cited (in its
nonlinear-PDE form, Gerdt 2008) by B\"achler et al.\ (2012). Note the exact
match with the involutive clause in the B\"achler et al.\ definition of a
differential simple system: \emph{``$S$ is (Janet) involutive if all
non-admissible prolongations in $(S_T)^=$ reduce to zero by $(S_T)^=$.''} Gerdt
also records \emph{Theorem 17}: an $L$-involutive basis of $[F]$ is a
\emph{differential Gr\"obner basis}; and \emph{Theorem 18}, an involutive
analogue of the Buchberger chain criterion --- the same $S$-polynomial /
$\Delta$-polynomial content that, on the RG side, \emph{is} coherence.

\section{The connecting theorem: passive $\Rightarrow$ coherent}

The bridge between the two vocabularies is a single, one-directional theorem,
stated and proved in Li--Wang and echoed in the Thomas literature.

\begin{quote}
\textbf{Li--Wang 1999, Theorem 3.} \emph{A passive d-asc set is coherent.}
\end{quote}

\noindent
The proof (Li--Wang \S4.1, ``Passivity and coherence'') runs through Wu's
tuple-decomposition and passivity theorems; the section's framing is explicit
that this direction is the useful one: ``\emph{coherent d-asc sets are, in some
sense, simpler than passive d-asc sets}.'' The converse is \emph{not} claimed.

\noindent
The same implication is stated operationally in the differential-Thomas paper:
B\"achler, Gerdt, Lange-Hegermann \& Robertz (2012), \S1: ``\emph{the output
systems are Janet-involutive $\dots$ and \textbf{hence they are coherent}},''
and, on the algorithm, ``\emph{the involutive approach, which we utilize to make
the subsystems coherent}.'' Direction: involutive $\Rightarrow$ coherent, via
Gerdt's involutivity criterion.

\paragraph{Correction to an earlier informal claim.} It is tempting (and was
asserted in passing during the discussion that produced this note) to attribute
a coherence $\Leftrightarrow$ involution \emph{equivalence} to W.~Seiler's
monograph \emph{Involution} (2010) and, worse, to his two open-access AAECC 2009
papers (``A combinatorial approach to involution and $\delta$-regularity, I \&
II''). This is not supported by those two papers: they are commutative-algebra
papers on involutive/Pommaret bases and $\delta$-regularity and contain
\emph{zero} occurrences of ``coherence,'' ``Rosenfeld,'' or ``regular
differential.'' They connect involution to Janet--Riquier \emph{passivity}, not
to RG coherence. The defensible, sourced statement is the \emph{implication}
above, not an equivalence, and its primary home is Li--Wang 1999 (Thm.~3) and
Gerdt 1999/2008, not Seiler. (Seiler's $\delta$-regularity work is nonetheless
the right reference for \emph{why} Pommaret-involution --- unlike Janet-involution
and unlike coherence --- is coordinate-dependent, which is a second, independent
reason the two notions cannot be identical.)

\section{The implication lattice}

Collecting Definitions 5, 6, 8 and Theorems 2, 3 of Li--Wang with Gerdt's
Theorem 16:
\[
\begin{gathered}
  \underbrace{\text{passive}}_{\substack{\text{Janet-involutive}\\ \text{Gerdt Def.~14/Thm.~16}}}
  \;\xRightarrow{\ \text{L--W Thm.~3}\ }\;
  \text{coherent}
  \\[1.5ex]
  \underbrace{\text{simple}}_{\text{L--W Def.~8}}
  \imp
  \underbrace{\text{regular}}_{\text{L--W Def.~6}}
  \;=\;\text{coherent}+\text{e-tri}
  \imp
  \text{coherent}.
\end{gathered}
\]

\begin{center}
\begin{tabular}{@{}p{2.3cm}p{5.6cm}p{3.0cm}p{2.7cm}@{}}
\toprule
Node & Defining condition & Decomposition & Source \\
\midrule
coherent & $\Delta$-polys reduce to $0$ & (property) & L--W Def.~5 \\
passive / involutive & non-mult.\ prolongations reduce to $0$ & (property) & Gerdt Def.~14, Thm.~16 \\
regular & coherent $+$ e-tri (sep./init.\ invertible on the component) & intersection (RG weak form) & L--W Def.~6 \\
simple (d-sim) & regular $+$ w-simple (triangular, non-vanishing init., square-free) & disjoint (Thomas) & L--W Def.~8 \\
\bottomrule
\end{tabular}
\end{center}

\noindent
Every arrow is \emph{strict} and \emph{one-directional}:
\begin{itemize}
  \item \textbf{coherent} is the base property --- it makes Rosenfeld's Lemma
        fire (differential $\to$ algebraic), and nothing more is assumed.
  \item \textbf{passive $\Rightarrow$ coherent} but not conversely: passivity is
        a specific, division-dependent (Janet) completion; coherence is the
        weaker, division-free, intrinsic property it implies (L--W Thm.~3).
  \item \textbf{regular $\Rightarrow$ coherent} but not conversely: regular adds
        the e-tri (whole-component invertibility of separants/initials) upgrade,
        which is what makes the saturated ideals radical (L--W Thm.~8).
  \item \textbf{simple $\Rightarrow$ regular} but not conversely: simple adds
        square-freeness / the w-simple conditions, which is what buys
        \emph{disjointness} of the Thomas decomposition and the characteristic-set
        property (L--W Cor.~2).
\end{itemize}

\section{Consequences for the RG-vs-Thomas comparison}

\begin{itemize}
  \item \textbf{Coherence is not the discriminator between RG and Thomas.} Both
        traditions require it (RG via Rosenfeld's Lemma; Thomas via the
        involutive clause that \emph{implies} it). The real discriminators are
        the \emph{extra} conditions: e-tri (regular) vs.\ e-tri $+$ square-free
        (simple), i.e.\ \emph{overlapping intersection} vs.\ \emph{disjoint
        partition}.
  \item \textbf{$\Delta$-polynomial $=$ differential $S$-polynomial.} B\"achler
        et al.\ (2012) say it outright (``the analogon of $s$-polynomials in
        differential algebra''), and Gerdt's Theorem 18 is the involutive
        Buchberger chain criterion. So the RG/Thomas relationship mirrors the
        algebraic Gr\"obner-basis / involutive-basis relationship: every
        involutive basis is a Gr\"obner basis but generally larger and more
        structured; likewise passive $\Rightarrow$ coherent, and simple is the
        more refined (disjoint) decomposition.
  \item \textbf{On ``is a differential Thomas decomposition always finer than a
        (parametric) RG decomposition?''} Fibrewise (same ranking, differential
        structure), yes: simple $\Rightarrow$ regular, and simple exposes the
        degenerate branches RG saturation folds away. But \emph{globally},
        including a parameter stratification, the two can become
        \emph{incomparable}, because they split along the discriminant loci that
        \emph{their own} orderings produce, and parametric RG carries a
        distinguished parameter layer (with deferred genericity conditions) that
        Thomas --- having ``no notion of parameter'' --- does not reproduce. The
        one-directional lattice here is what holds \emph{after} fixing a common
        ambient ring and ranking; it is not a claim of global refinement.
\end{itemize}

\section{Primary sources}

\begin{itemize}
  \item Z.~Li \& D.~Wang, \emph{Coherent, regular and simple systems in zero
        decompositions of partial differential systems}, Systems Science and
        Mathematical Sciences \textbf{12} (Suppl.), May 1999, 43--60.
        [Defs.~5 (coherent), 6 (regular), 8 (d-sim); Thm.~2 (generalized
        Rosenfeld), Thm.~3 (passive $\Rightarrow$ coherent), Thm.~8 (radicality),
        Thm.~9/Cor.~2 (simple).] Local archive:
        \nolinkurl{~/project/papers/li-wang-1999-coherent-regular-simple-systems-zero-decompositions-pde.pdf}
        (scanned; page images read directly).
  \item V.~P.~Gerdt, \emph{Completion of Linear Differential Systems to
        Involution}, CASC 1999, 115--137 (arXiv:math/9909114). [Def.~14
        (involutive), Thm.~16 (local involutivity criterion: non-multiplicative
        prolongations reduce to $0$), Thm.~17 (involutive $\Rightarrow$
        differential Gr\"obner), Thm.~18 (involutive Buchberger chain criterion);
        prose: passive $=$ Janet-involutive.] Local archive:
        \nolinkurl{~/project/papers/gerdt-1999-completion-linear-differential-systems-to-involution.pdf}.
  \item V.~P.~Gerdt, \emph{On Decomposition of Algebraic PDE Systems into Simple
        Subsystems}, Acta Appl.\ Math.\ \textbf{101} (2008), 39--51. [The
        involutivity criterion in nonlinear-PDE form, as cited by B\"achler et
        al.]
  \item T.~B\"achler, V.~Gerdt, M.~Lange-Hegermann \& D.~Robertz, \emph{Thomas
        Decomposition of Algebraic and Differential Systems}, J.~Symbolic
        Comput.\ \textbf{47} (2012), 1233--1266 (arXiv:1008.3767). [Def.~2.1
        (algebraic simple), Def.~3.3 (differential simple $=$ algebraically
        simple $+$ involutive $+$ minimal $+$ inequation-reduced); ``Janet-involutive
        $\dots$ hence coherent''; $\Delta$-poly $=$ $s$-poly analogue.] Local
        archives (published + preprint) in \nolinkurl{~/project/papers/}.
  \item W.~M.~Seiler, \emph{A Combinatorial Approach to Involution and
        $\delta$-Regularity, I \& II}, AAECC \textbf{20} (2009). [Involutive /
        Pommaret bases, $\delta$-regularity. \emph{Not} a source for a
        coherence $\Leftrightarrow$ involution equivalence --- see \S4
        correction. Relevant for the coordinate-dependence of Pommaret
        involution.]
\end{itemize}

\bigskip
\noindent\emph{Note prepared as computational/literature backing for
NewMethod; reconciles the terminology used across the RG and differential-Thomas
ports (\code{joca.sage}, \code{joca-thomas-native.sage}, and the parametric-RG
port).}

\end{document}
