
\documentclass{article}

\title{An Enhanced Gr\"obner Factorizer for\\ solving systems of Polynomial Equations}
\author{Brent Baccala}

\usepackage{amsmath}
\usepackage{amsfonts}

% for \url
\usepackage[hidelinks]{hyperref}

% for \setmarginsrb
\usepackage{vmargin}

\setmarginsrb{0.5in}{1in}{0.5in}{0.5in}{20pt}{20pt}{20pt}{20pt}

\begin{document}
\parindent 0pt

\maketitle

\begin{abstract}
Factoring polynomials to simplify Gr\"obner basis calculations has
been state-of-the-art for over 30 years.
The author has developed an extension to previous factorized Gr\"obner basis algorithms
that uses monotone dualization combined with some simple simplifications, and
implemented the algorithm using the Sage computer algebra system.
\end{abstract}

%%\section*{The Algorithm}

\parskip 12pt

\subsection*{Previous Work}

\cite{Grabe94} ---

\cite{Grabe96} --- continues study of factorized Gr\"obner bases from \cite{Grabe94} and \cite{Grabe95}.
The introduction credits Seidenberg's work in the 1970s and 1980s, leading up to the GTZ
algorithm \cite{GTZ}, ``the first general primary decomposition algorithm''

\cite{Grabe06} --- Linz talk.  Cited by Hans Sch\"onemann. ``Among the methods to split polynomial systems into smaller pieces probably the Gr\"obner
factorizer method attracted the most theoretical attention'', citing Czapor (\cite{Czapor87} and \cite{Czapor89})

Geddes, Czapor, Labahn wrote ``Algorithms for Computer Algebra''.  All were involved with developing MAPLE.
Geddes and Gonnet were professors at the University of Waterloo when they founded Maplesoft.
Czapor was a Ph.D. student of Geddes.

%% \subsection*{Introduction}

Hans-Gert Gr\"abe, in \cite{Grabe94}, proposed a modified Buchberger algorithm
that aggressively factors the polynomials appearing in a Gr\"obner basis
calculation in order to split the original polynomial system into simpler subsystems.
Gr\"abe implemented his ideas in REDUCE package called CALI, which has since been ported
to Singular and is available as the Singular function {\tt facstd}.

Gr\"abe algorithm:

\begin{itemize}
\item try to factor each polynomial in a preprocessing step
\item during the Buchberger algorithm, try to factor each new reduced S-polynomial
\item when finished, apply tail reduction and check for factorization again
\end{itemize}

Singular \cite{DGPS}, in its {\tt primdec} library for primary decomposition,
adds a preprocessing step to this algorithm:

\begin{itemize}
\item check for simple reductions in a preprocessing step
\item use the factoring Gr\"obner basis algorithm (with an option to disable the factorization) to complete the GTZ algorithm.
\end{itemize}

The ``simple reductions'' of the Singular algorithm (implemented in the {\tt simplifyIdeal} procedure)
are formed from polynomials in which
there is a variable that appears only once in the polynomial, and only as an isolated term with degree one.
Thus, $x$, $x-1$, and $x+y^2$ are all candidates (in $x$) for simple reduction, and
the reductions would be to substitute $x=0$, $x=1$, or $x=-y^2$ (respectively) into all
polynomials in the system.

These ``simple reductions'' are equivalent to reordering the variables
so that the most significant variable has a polynomial in
the system with a term consisting of that variable only,
then using this polynomial to apply a tail reduction step to the remaining polynomials.

The check for ``simple reductions'' is only done once, as a preprocessing step,
and not attempted again after the Gr\"obner basis calculation has begun.
Combined with the factoring Gr\"obner basis algorithm introduced by Gr\"abe,
the algorithm is seen to consist of attempting simple reductions, then
attempting factorization as the first step in the Gr\"obner basis algorithm.

I've found that the first factoring step often yields systems of equations in
which simple reductions can be found that were not present prior to
the factorization, so it seems profitable to check for further simple reductions.

My algorithm:

\begin{itemize}
\item check for simple reductions and apply them
\item try to split the system by factoring the polynomials
\item repeat these two steps they yield no further reductions
\item use the Gr\"abe algorithm to apply the GTZ algorithm
\end{itemize}

The approach to the factorized Gr\"obner basis algorithm in \cite{Grabe06}
is described like this ($B$ is the system of polynomials to be solved):

\begin{quote}
During a preprocessing interreduce $B$ and try to factor each polynomial $f \in B$. If $f$ factors,
replace $B$ by a set of new problems, one for each factor of $f$. Update the side conditions
and apply the preprocessing recursively. This ends up with a list of interreduced problems
with non factoring base elements.
\end{quote}

This suggests that the factorization and splitting is performed independently on each
polynomial in the system.  If the system is large and many of the polynomials factor,
then this can lead to exponential growth in the number of systems.

This exponential growth in the number of systems can sometimes be mitigated
by first factoring all of the polynomials in the system, then performing
a logical transformation, as follows.

Consider a polynomial $f$ that factors as $f_1\cdot f_2\cdot f_3$.
The condition $f=0$ is then equivalent to the logical system ($f_1=0 \vee f_2=0 \vee f_3=0$).
If another polynomial $g$ factors as $g_1 \cdot g_2$, then the simultaneous
systems of equations $f=0; g=0$ is equivalent to this logical system:

\begin{equation}
(f_1=0 \vee f_2=0 \vee f_3=0) \wedge (g_1 = 0 \vee g_2 = 0)
\end{equation}

This logical system is in conjunctive normal form (CNF) and can be
converted to an equivalent system in disjunctive normal form (DNF), specifically:

\begin{equation}
\begin{split}
(f_1=0 \wedge g_1 = 0) \vee (f_2=0 \wedge g_1 = 0) \vee (f_3=0 \wedge g_1 = 0) \\
\vee (f_1=0 \wedge g_2 = 0) \vee (f_2=0 \wedge g_2 = 0) \vee (f_3=0 \wedge g_2 = 0)
\end{split}
\end{equation}

In this example, we have now obtained six simultaneous systems of polynomial equations,
each composed of irreducible polynomials.  A solution to any of them will solve the
original system.  The alegbraic variety defined by the original system is the union
of the six irreducible algebraic varieties defined by the factored systems.

Therefore, I do not split the system after each factorization, as in \cite{Grabe06},
but rather factor all of the polynomials in the system, then consider
each factor as a logical variable that is either true or false, depending on
whether the factor is or is not equal to zero.  Form a logical system in CNF
from the original polynomials and convert it to DNF to obtain systems of polynomials
formed from the irreducible factors.

For physically realistic problems, we hope that there is enough overlap between
the factorizations of the various polynomials to make this conversion desirable.

For example, one of the systems I'm concerned with (ansatz 16.6 for helium's ground state)
has 3486 polynomials.  2341 of them are irreducible; 969 of them factor into two irreducible factors, and 176 of them
factor into three irreducible factors.  Splitting on each of these factors independently,
the Gr\"abe algorithm would produce $2^{969} 3^{176} > 10^{375}$
distinct systems.  Yet there are only 2950 distinct factors, not $2341+2\cdot969+3\cdot176=4807$,
so there's a lot of overlap between the factors of the various polynomials.
Using CNF-to-DNF conversion, we obtain not $2^{969} 3^{176}$ systems, but only 48.

An obvious question is whether every Boolean function in CNF has a unique DNF.  In general, the
answer is no, but the specific Boolean functions defined by polynomial systems are {\it monotone} Boolean functions,
because they do not involve the logical inverse of any logical variable, i.e,
they depend only on $f=0$ and not $f\ne 0$.  Monotone Boolean functions admit
a single unique DNF of unique prime implicants, due to a theorem stating that for monotone Boolean functions,
all prime implicants are essential.  Therefore, up to ordering, the DNF is unique.

Also, a logical expression in DNF can be inverted to obtain a logical expression in CNF
for the inverse Boolean function.  Since the inverse of a monotone Boolean function is also
a monotone Boolean function, after a CNF-to-DNF conversion is performed on the inverse function,
the inverse function's DNF can be inverted to obtain the original CNF.  This procedure
is referred to in the literature as ``monotone dualization'', and is a very convenient
property for testing the correct operation of a software implementation.

\section*{Implementation}

I've implemented these ideas using the Sage computer algebra system.

Sage's basic design philosophy is to use existing programs and libraries
whenever practical, interconnecting them all using the Python language,
with a Python interpreter being Sage's primary user interface.
Cython is used heavily.
Singular is used as a backend for many polynomial calculations,
including the GTZ algorithm which is implemented using Singular's
native Gr\"obner basis algorithms, including a factorized
Gr\"obner basis algorithm, used by default in a GTZ calculation.


My implementation of the algorithm proposed above has several components:

\begin{itemize}
\item A Python library that implements the overall algorithm
\item A standard (i.e, unmodified) PostgreSQL database that stores the intermediate calculations
\item A multithreaded C++ program to perform the monotone dualization
\item Cython code to implement the simplification logic
\item Cython code to implement a faster pickling algorithm for polynomials over finite fields
\end{itemize}

The core process of the algorithm is to extract a polynomial subsystem from the SQL database,
apply any simplifications, factor the polynomials, compute the monotone dualization,
then insert the resulting subsystems back into the SQL database, each flagged for further processing.

Parallel operation of the algorithm is fairly simple --- any number of processes on any
number of nodes can be simultaneously processing different subsystems.  PostgreSQL
supports both transactions and network operations, making it easy to support concurrency
over a cluster.

Prior implementations of the factorized Gr\"obner basis algorithm
store the polynomial systems in RAM.  Because I expect a large number
of systems to split off (on the order of millions), I store the
systems in a SQL database, using Python's pickling format.

\vfill\eject
\begin{thebibliography}{9}

\bibitem{Czapor87}
S. R. Czapor. Solving algebraic equations via Buchberger’s algorithm. In Proc. EUROCAL’87,
volume 378 of Lect. Notes Comp. Sci., pages 260 -- 269, 1987.

\bibitem{Czapor89}
S. R. Czapor. Solving algebraic equations : Combining Buchberger’s algorithm with
multivariate factorization. J. Symb. Comp., 7:49 -- 53, 1989.

\bibitem{Grabe94}
Hans-Gert Gr\"abe (1994). ``On Factorized Gr\"obner Bases''

\url{https://d-nb.info/1249927226/34}

\bibitem{Grabe95}
Hans-Gert Gr\"abe (1995). ``Triangular systems and factorized Gr\"obner bases''

\bibitem{Grabe96}
Hans-Gert Gr\"abe (1996). ``Minimal Primary Decomposition and Factorized Gr\"obner Bases''

\bibitem{Grabe06}
Gr\"abe (2006).  ``The Groebner Factorizer and Polynomial System Solving'',
Talk given at the Special Semester on Groebner Bases, Linz 2006.

\url{https://www.ricam.oeaw.ac.at/specsem/srs/groeb/download/06_02_Solver.pdf}

\bibitem{GTZ}
Gianni, P., Trager, B., Zacharias, G.: ``Gr\"obner bases and primary decomposition of polynomial ideals''.
{\it J. Symb. Comp.} 6 (1988), 149 - 167.

\bibitem[DGPS]{DGPS}
Decker, W.; Greuel, G.-M.; Pfister, G.; Sch{\"o}nemann, H.:
\newblock {\sc Singular} {4-4-0} --- {A} computer algebra system for polynomial computations.
\newblock {https://www.singular.uni-kl.de} (2024).

\end{thebibliography}

\end{document}
